---
title: "R for SPSS users"
author: "Chris Beeley"
date: "07/05/2019"
output: ioslides_presentation
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

```

## Introduction

* RStudio
* Loading and saving data
* Common statistical tests
* Summarising data
* Other data operations (e.g. filtering)

## RStudio

* Panes
* The terminal
* Code files and running lines of code
* Files
* Environment variables

## Packages

* R has support for packages, which enhance and add to the way R works
* There are over 14,000 packages
* Some are very useful to many, many are for very specific tasks, such as analysing eye tracking data or MRIs

## The tidyverse

* I'm going to teach you R the tidyverse way
* As opposed to the base R way
* "The tidyverse is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures"
* We'll talk more about the tidyverse as the day progresses

## Reading data

* read_csv
    - From the readr package, which is part of the tidyverse
* You can point and click within RStudio to load in data with read_csv
* This will also generate the code that was used to load in the data
* This can then be copied into the script

## Code

<!-- https://www.reddit.com/r/datasets/comments/1uyd0t/200000_jeopardy_questions_in_a_json_file/ -->

```{r}

library(readr)
JEOPARDY <- read_csv("JEOPARDY.csv")

# note that I added the correct column for teaching purposes!

```

EXERCISE ONE

## Dataframes and data checking

* A dataframe is a load of rows and columns, with each column containing the same data type (just like in SPSS)
* You can check your dataframe in the following ways
    - Click in the environment panel
    - str()
    - head() + tail()
    - summary()
    
## Dataframes and data checking

* You can access each column on its own using $ or [[""]]
    - df$PatientID
    - df["PatientID"]
    - str(), head(), tail(), summary() all still work
    - table() can be useful
    - Can use [] to subset df$PatientID[6]
    - Or multiple df$PatientID[10 : 20]
    
## Data types

* Numbers
* Characters
* Dates
* Boolean
* Factors

## Let's make a factor

```{r}

JEOPARDY$Round <- factor(JEOPARDY$Round, 
                         levels = c("Jeopardy!", 
                                    "Double Jeopardy!", 
                                    "Final Jeopardy!"))

```

EXERCISE TWO

## Missing data and other strange values

* NA
* Inf
* NaN
* NULL

## Descriptive statistics

```{r}

library(dplyr)
library(ggplot2)

NHANES <- read_csv("NHANES.csv")

adults <- NHANES %>% 
    filter(Age >= 18)

# numerical data

mean(adults$Weight)

mean(adults$Weight, na.rm = TRUE)

```

## Descriptive statistics

```{r}

sd(adults$Weight, na.rm = TRUE)

median(adults$Weight, na.rm = TRUE)

min(adults$Weight, na.rm = TRUE)

max(adults$Weight, na.rm = TRUE)

```

## Summarising a variable

```{r}

summary(adults$Weight, na.rm = TRUE)

```

This command works on lots of R objects, not just variables (we already saw its being used with a dataframe).

## Describing categorical data

```{r}

# categorical data

unique(adults$Race1)

table(adults$Race1)

```

## Percentages

```{r}

prop.table(table(adults$Race1)) * 100

```

EXERCISE THREE

* mean, median, min, max, sd, summary, unique, table, prop.table

## Correlation

```{r}

# standard

cor.test(adults$Height, adults$Weight)

```

## Non parametric

```{r}

cor.test(adults$Height, adults$Weight, method = "spearman")

```

## Non parametric

```{r}

cor.test(adults$Height, adults$Weight, method = "kendall")

```

## Correlation plots

```{r, eval = FALSE}

library(corrplot)

data_corr <- cor(
  adults %>% 
    select(Age, Poverty, HomeRooms, Weight, Height, 
           Pulse, DirectChol, AlcoholYear),
  use = "pairwise.complete.obs")

corrplot(data_corr, method = "color")

```

## Correlation plots

```{r, echo = FALSE}

library(corrplot)

data_corr <- cor(
  adults %>% 
    select(Age, Poverty, HomeRooms, Weight, Height, 
           Pulse, DirectChol, AlcoholYear),
  use = "pairwise.complete.obs")

corrplot(data_corr, method = "color")

```

## t test

```{r}

t.test(Weight ~ Gender, data = NHANES)

```

## ANOVA

```{r}

library(dplyr)

PlantGrowth %>% 
  group_by(group) %>%
  summarise(
    count = n(),
    mean = mean(weight, na.rm = TRUE),
    sd = sd(weight, na.rm = TRUE)
  )

```

## ANOVA

```{r}

# Compute the analysis of variance
# beware if you have an unbalanced design, things get a little more complicated

res.aov <- aov(weight ~ group, data = PlantGrowth)

# Summary of the analysis
summary(res.aov)

```

## Non parametric difference of means

```{r}

kruskal.test(weight ~ group, data = PlantGrowth)

```

## Regression

* Using the National Health and Nutrition Examination Survey (NHANES) data from the CDC
* Do some plots
* Fit a multiple regression
* Produce diagnostic plots

## Histogram

```{r}

adults %>% 
    ggplot(aes(x = Weight)) + geom_histogram()

```

## Boxplots

```{r}

adults %>% 
    ggplot(aes(x = Gender, y = Weight)) + geom_boxplot()

```

## Add a violin plot

```{r}

adults %>% 
    ggplot(aes(x = Gender, y = Weight)) + geom_boxplot() + 
    geom_violin(alpha = .3)

```

## Fit a regression model

```{r}

model1 <- lm(Weight ~ Age + Gender + Height + Poverty + 
                 AlcoholDay + PhysActiveDays, data = adults)

summary(model1)

```

## Results

```{r}

adults %>% 
    ggplot(aes(x = Height, y = Weight)) + geom_point() + 
    geom_smooth(method = "lm") + facet_wrap(~ Gender)

```

## Diagnostic plots

```{r}

par(mfrow = c(2, 2))

plot(model1)

```

EXERCISE 4

Histogram:



## Crosstabs and chi square

```{r}

library(readxl)

titanic <- read_excel("titanic3.xls")

table(titanic$survived, titanic$pclass)

prop.table(table(titanic$survived, titanic$pclass), 2) * 100

```

## Crosstabs

```{r}

titanic_table <- xtabs(~ pclass + survived + sex, data = titanic)

plot(titanic_table)

```

## Chi square

```{r}

chisq.test(table(titanic$survived, titanic$pclass))

```

## GLM

```{r}

glm_model <- glm(Correct ~ Value, data = JEOPARDY, family = binomial)

summary(glm_model)

```

## GLM

```{r}

predict(glm_model, 
        newdata = data.frame(Value = c(200, 400, 600, 800, 1000)), 
        type = "response")

```

## Survival analysis

```{r}

library(survival)

fit <- survfit(Surv(time, status) ~ sex, data = lung)
fit

summary(fit)$table

```

## Plotting

```{r}

library(survminer)

ggsurvplot(fit, pval = TRUE, conf.int = TRUE, risk.table = TRUE, 
          risk.table.col = "strata", linetype = "strata", 
          surv.median.line = "hv", ggtheme = theme_bw(), 
          palette = c("#E7B800", "#2E9FDF"))

```

## Mixed effects models

```{r}

library(lme4)

ggplot(sleepstudy, aes(x = Days, y = Reaction, col = Subject)) + 
    geom_point() + stat_smooth(method= 'lm', se = F) + 
    facet_wrap(~ Subject) + guides(col = F)

```

## Random intercepts

```{r}

lmemodel1 <- lmer(Reaction ~ Days + (1 | Subject),
            data = sleepstudy)

summary(lmemodel1)

```

## Random slopes
 
```{r}

lmemodel2 <- lmer(Reaction ~ Days + (0 + Days | Subject),
            data = sleepstudy)

summary(lmemodel2)

```

## Random slopes and random intercepts

```{r}

lmemodel3 <- lmer(Reaction ~ Days + (1 + Days | Subject),
            data = sleepstudy)

summary(lmemodel3)

```

## The tidyverse

* The tidyverse is a set of packages which will help you with your data
    - ggplot2
    - dplyr
    - tidyr
    - readr
    - purrr
    - tibble
    - stringr
    - forcats
    
## Tidy data

* Each variable forms a column
* Each observation forms a row

```{r}

medals = data.frame(
  country = c("Norway", "Germany", "Canada"),
  gold = c(14, 14, 11), 
  silver = c(14, 10, 8),
  bronze = c(11, 7, 10)
)

medals

```

## Tidying

```{r}

library(tidyr)

gather(medals, key = Type, value = Medals, -country)

```

    
## ggplot2

* Accepts only tidy data
* Each is made from an aes() and a geom_...()

```{r, eval = FALSE}

housing <- read_csv("landdata-states.csv")

subset_housing <- housing %>% 
  filter(State %in% c("MA", "TX"))

ggplot(subset_housing,
       aes(x=Date,
           y=Home.Value,
           color=State))+
  geom_line()

```

## ggplot2

```{r, echo = FALSE}

housing <- read_csv("landdata-states.csv")

ggplot(filter(housing, State %in% c("MA", "TX")),
       aes(x=Date,
           y=Home.Value,
           color=State))+
  geom_point()

```

## A diversion about pipes

* Pipes are ubiquitous in R code in the wild now
* You will need to understand them to read code that you find online
* Basically, mpg %>% summary() is equivalent to summary(mpg)
* gapminder %>% filter(year > 1960) is equivalent to filter(gapminder, year > 1960)
* In general, x %>% f %>% g %>% h is equivalent to h(g(f(x)))

## A larger example

```{r, eval = FALSE}

gapminder %>%
    filter(year > 1960) %>%
    group_by(continent, year) %>%
    summarise(meanLife = mean(lifeExp))

```

Is equivalent to

```{r, eval = FALSE}

summarise(
    group_by(
        filter(gapminder, year > 1960),
        continent, year),
    meanLife = mean(lifeExp))

```

## Using facets

```{r}

ggplot(housing, aes(x = Date, y = Home.Value, color = State)) + 
  geom_line()

```

## Nicer plot

```{r}

ggplot(housing, aes(x = Date, y = Home.Value)) + geom_line() +
   facet_wrap(~State, ncol = 10)

```

## Loading SPSS data

* This could be painful the first time!
    - I can help
* Once you've got it loaded resave:
    - As csv
    - As .Rdata
    
```{r}

library(haven)

dummy <- read_spss("Dummy_SPSS.sav")

labelled_dummy <- as_factor(dummy)

```

## Tidyverse- mutate

```{r}

dummy <- dummy %>% 
  mutate(Discharge_Location = na_if(MHA_classification, 999))

```

## Tidyverse- group_by

```{r}

labelled_dummy %>% 
  group_by(Ethnicity) %>% 
  count()

```

## Tidyverse- filter

```{r}

labelled_dummy %>% 
  filter(Previous_convictions > 10) %>% 
  summarise(mean(T1_HCR20_Hist, na.rm = TRUE))

```

## Saving data

```{r, eval = FALSE}

# .Rdata

save(dummy, labelled_dummy, file = "spss_dummy.Rdata")

# .RDS

saveRDS(dummy, file = "spss_dummy_unlabelled.rds")

# they load differently

load("spss_dummy.Rdata")

dummy_new <- readRDS("spss_dummy_unlabelled.rds")

```

## A note on packages

* You will need to install the packages from today on your machine
* You can do this with install.packages() or by pointing and clicking with RStudio
* You can install several (such as tidyr, ggplot2, and readr) all at once with install.packages("tidyverse")

EXERCISE 5
